INGRID: Creating Languages in MPS from ANTLR Grammars

### 2016-10-30 chapters/language_definition.tex
remarks - first part

@@

As we have already shown in previous sections, it is very easy to write a grammar in a way that will cause problems during import into MPS (or once it is imported).
Problems might occur during the creation of any aspect.
Here we will look at some general problems/issues that grammar import poses (might pose).
We will talk about these problems with respect to the challenges/obstacles we tried to overcome and which we described in previous sections (dedicated to structure, editor, and texgen, respectively).
We also show a few examples of additional possible complications.

\paragraph{Adjusting grammars.}
There are some cases, where altering the input grammar might yield far better MPS language.
We will show two examples how the usability of the resulting MPS language can be improved.
Let us look at the definition of an XML attribute in Figure~\ref{fig:xmlattribute}.

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
	\antlrlexerrule{STRING} : \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
	       | \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'} ;
\end{alltt}
\end{framed}
\caption{Definition of an XML attribute}
\label{fig:xmlattribute}
\end{figure}

The original XML grammar has quotes as a part of the value.
For the resulting MPS language, it would mean that there would be a placeholder for the attribute value that would expect us to input the leading and trailing quote together with the value too each time.
It would also be marked red (by the syntax checker) unless we enter both quotes inside the value since the regular expression checking for quotes will not match.
The user might be confused by this and will not be able to tell why his string value is incorrect.

In our SimpleXML language, we adjusted the grammar easily in the manner that we show in Figure~\ref{fig:xmladjustgrammar}.

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
	       | \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'} ;
	\antlrlexerrule{TEXT1} : \antlrregex{~["]*} ;
	\antlrlexerrule{TEXT2} : \antlrregex{~[']*} ;
\end{alltt}
\end{framed}
\caption{Adjusted grammar of SimpleXML}
\label{fig:xmladjustgrammar}
\end{figure}

We turned quotes into literals, thus ensuring that they will only appear in the projectional editor as fixed constant cells.
We will not have to encapsulate the value in them each time.
The user will only have to choose, which attribute version he wants to use (single or double quotes).

As the second example, we will use the ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4} language otherwise known as JavaScript.
Every statement in JavaScript needs to be either followed by a semicolon, newline, file end or end of the block --- see the Figure~\ref{fig:javascriptstmt}.

\begin{figure*}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{eos} : \antlrlexerrule{SemiColon} | \antlrlexerrule{EOF} | {lineTerminatorAhead()}? 
	    | {{\_}input.LT(1).getType() == \antlrlexerrule{CloseBrace}}? ;
	\textcolor{gray}{// Example reference of the eos rule}
	\antlrparserrule{breakStmt} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos} ;
\end{alltt}
\end{framed}
\caption{Statements in JavaScript}
\label{fig:javascriptstmt}
\end{figure*}

Because there are multiple options, our import plugin would create a placeholder at the end of every statement.
Every language element representing a statement would have one child of the \mpsinterface{IEos} interface type.
This placeholder needs to be manually filled in for each statement.

Since the projectional editor has much bigger power over the form of the code, we (a user) might want to have each statement on a separate line.
Since we can differentiate between statements on the AST level, we do not need an explicit separator between them.
This means that we might want to simplify the language and leave the semicolon out, or leave it just as a constant fixed part of the projectional editor, but not as something the user must explicitly fill in.
Then we can just put each statement on a separate line as it is usual for JavaScript code, but we do not need the semicolon anymore.
This small adjustment is very quick when done inside the grammar.
We just change the \antlrparserrule{eos} rule to the form shown in Figure~\ref{fig:eosrule}.

\begin{figure}[ht]
\centering
\begin{framed}
\begin{alltt}
	\antlrparserrule{eos} : \antlrliteral{;} ;
\end{alltt}
\end{framed}
\caption{eos parser rule}
\label{fig:eosrule}
\end{figure}

This way, we can change the grammar in a way that it will not describe the same ECMAScript language as before, but will definitely make our MPS language more usable.
TextGen aspect may be defined in such a way that puts the semicolon back (after each statement) in the generated plain-text representation.
This would be very important especially when JavaScript is available as another base language (which is actually a plan (future work) of the JetBrains company).

Above, we have shown that adjusting the input grammar might be a very fast mean of tweaking/improving the usability of generated MPS language, and sometimes it may be the only proper solution in some complex situations.
In general, the question we can ask ourselves is whether one aims for a full precise/faithful port of the input language or for an MPS alternative.

>>>>

During our work on the INGRID method, we have also observed that practical usability of a resulting MPS language depends on the specific manner in which the input ANTLR grammar is defined.
We discuss several issues and our workarounds in this section.

\paragraph{Adjusting grammars.}
In some cases, a small adjustment of the input grammar before the run of INGRID might yield a better and more useful MPS language.
Here we illustrate this on two simple examples.

For the first example, we use the definition of an XML attribute in Figure~\ref{fig:XMLATTRIB}, which is taken from the original XML grammar.
The lexer rule \antlrlexerrule{STRING} actually says that quotes make a part of the attribute's value.
If the MPS language would faithfully match the grammar, the user would have to always input the leading and trailing quote together with the value in the projectional editor.

\begin{figure}[ht]
\begin{framed}
\begin{alltt}
  \antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{=} \antlrlexerrule{STRING} ;
  \antlrlexerrule{STRING} : \antlrliteral{"} \antlrregex{~["]*} \antlrliteral{"}
         | \antlrliteral{\textbackslash'} \antlrregex{~[']*} \antlrliteral{\textbackslash'} ;
\end{alltt}
\end{framed}
\caption{Definition of an XML attribute}
\label{fig:XMLATTRIB}
\end{figure}

In our SimpleXML language, we adjusted the grammar easily in a way that we show in Figure~\ref{fig:XMLADJUST}.
We turned quotes into literals, ensuring that they will only appear in the projectional editor as string constant cells.

\begin{figure}[ht]
\begin{framed}
\begin{alltt}
  \antlrparserrule{attrib} : \antlrlexerrule{Name} \antlrliteral{="} \antlrlexerrule{TEXT1} \antlrliteral{"}
         | \antlrlexerrule{Name} \antlrliteral{=\textbackslash'} \antlrlexerrule{TEXT2} \antlrliteral{\textbackslash'} ;
  \antlrlexerrule{TEXT1} : \antlrregex{~["]*} ;
  \antlrlexerrule{TEXT2} : \antlrregex{~[']*} ;
\end{alltt}
\end{framed}
\caption{Adjusted fragment of the SimpleXML grammar}
\label{fig:XMLADJUST}
\end{figure}

The second example relates to a fragment of the JavaScript language, also known as ECMAScript\footnote{https://github.com/antlr/grammars-v4/blob/master/ecmascript/ECMAScript.g4}.
Every statement in JavaScript needs to be terminated by a semicolon, newline, end of the file, or end of the block --- see the Figure~\ref{fig:JAVASCRIPTSTMT}, which contains the corresponding fragment of the ECMAScript grammar.
Since there are multiple options, a user would have to select one of them for each statement in the editor.
Technically, every language concept representing a statement would contain one child node of the interface type \mpsinterface{IEos}, which has to be assigned a proper object (i.e., an AST node corresponding to one of the options listed above).

\begin{figure}[ht]
\begin{framed}
\begin{alltt}
\small
\antlrparserrule{eos} : \antlrlexerrule{SemiColon} | \antlrlexerrule{EOF} | {lineBreakAhead()}? 
    | {{\_}input.LT(1).type() == \antlrlexerrule{CloseBrace}}? ;
\textcolor{gray}{// example reference to the eos rule}
\antlrparserrule{breakStmt} : \antlrlexerrule{Break} \antlrlexerrule{Identifier}? \antlrparserrule{eos} ;
\end{alltt}
\end{framed}
\caption{Statements in JavaScript}
\label{fig:JAVASCRIPTSTMT}
\end{figure}

Since MPS can differentiate between statements on the AST level, there is no need for an explicit separator.
For example, we can put each statement on a distinct line, as is usual for JavaScript code, and keep just the semicolon as a fixed literal in the projectional editor.
The \antlrparserrule{eos} rule must be changed to this form: \antlrparserrule{eos} : \antlrliteral{;}.
A small adjustment of the grammar, like this one, is a very quick solution and definitely makes the generated language more usable in MPS.
On the other hand, this adjustment of the original ANTLR grammar, performed just for the purpose of creating the MPS language, changes the grammar in such a way that it does not precisely describe the standard JavaScript language.
To ensure compatibility with JavaScript, the TextGen aspect in the MPS language should put the semicolon after each statement in the plain-text representation of a program code.

@@
